Unlocking the Potential of microRNAs: Machine Learning Identifies Key Biomarkers for Myocardial Infarction Diagnosis

### 2023-08-07 Revised Version/Article/sn-article.tex
The version after revision

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{tabularx}%
%%%%

@@

\author[2, 3]{\fnm{Reza} \sur{Rahbarghazi}}

\author[4]{\fnm{Zeynab} \sur{Piryaei}}

\author[5]{\fnm{Mahdad} \sur{Esmaeili}}

\author[6]{\fnm{Çığır} \sur{Biray Avcı}}

@@

\author[4*]{\fnm{Kaveh} \sur{Kavousi}}\email{kkavousi@ut.ac.ir}

\affil[1]{\orgdiv{Department of Medical Nanotechnology}, \orgname{Faculty of Advanced Medical Sciences}, \orgaddress{\street{Tabriz University of Medical Sciences}, \city{Tabriz}, \country{Iran}}}

\affil[2]{\orgdiv{Stem Cell Research Center}, \orgaddress{\street{Tabriz University of Medical Sciences}, \city{Tabriz}, \country{Iran}}}

\affil[3]{\orgdiv{Department of Applied Cell Sciences}, \orgname{Faculty of Advanced Medical Sciences}, \orgaddress{\street{Tabriz University of Medical Sciences}, \city{Tabriz}, \country{Iran}}}

\affil[4]{\orgdiv{Laboratory of Complex Biological Systems and Bioinformatics (CBB), Department of Bioinformatics}, \orgname{Institute of Biochemistry and Biophysics (IBB)}, \orgaddress{\street{University of Tehran}, \city{Tehran}, \country{Iran}}}

\affil[5]{\orgdiv{Medical Bioengineering Department}, \orgname{Faculty of Advanced Medical Sciences}, \orgaddress{\street{Tabriz University of Medical Sciences}, \city{Tabriz}, \country{Iran}}}

\affil[6]{\orgdiv{Medical Biology Department}, \orgname{School of Medicine}, \orgaddress{\street{Ege University}, \city{İzmir}, \country{Türkiye}}}

@@

\textbf{Results:} We identified hsa-miR-21-3p, hsa-miR-186-5p, and hsa-miR-32-3p as the
differentially expressed miRNAs, and a set including hsa-miR-186-5p, hsa-miR-21-3p,
hsa-miR-197-5p, hsa-miR-29a-5p, and hsa-miR-296-5p as the optimum set of miRNAs selected by
their AUC-ROC. Both biomarker sets could distinguish healthy from
not-healthy samples with complete accuracy. The best performance for the
classification of CAD and MI was achieved with an SVM model trained
using the biomarker set selected by AUC-ROC, with an AUC-ROC of 0.96 and
an accuracy of 0.94 on the test data.

@@

Over the last 50 years, numerous attempts have been made to use
biomarkers to facilitate diagnosis, assess the risk, follow-up therapy, and determine therapeutic efficacy in CVD candidates. Based on released
guidelines, cardiac troponins (cTns) are used as a highly sensitive and
accurate approach for detecting MI. Despite these inherent advantages,
the high sensitivity of cTn-based assays has also led to more
false-positive results \cite{17}, necessitating the advent
and development of new modalities with pathological value. To improve
the diagnostic value of existing MI biomarkers, a combination of
complementary biological markers, such as microRNAs (miRNAs) and other
genetic factors, has been proposed. Previous research supports the
notion that miRNAs exhibit great potential as alternative biomarkers for
CVD detection and follow-up \cite{miR}. It has been suggested
that miRNAs possess 18-22 nucleotides and play a crucial role in the
regulation of gene expression. Evidence indicates that miRNAs are
involved in the pathogenesis of cardiac tissue injury \cite{m1}. Several biological processes, such as angiogenesis, cardiomyocyte growth and contractility, lipid metabolism, plaque
formation, and cardiac rhythm, are regulated by miRNAs \cite{m2}. 
% Change S
Circulating and tissue-specific miRNAs have shown promise as diagnostic and prognostic biomarkers across a range of cardiovascular diseases, including MI and other conditions such as CAD, heart failure, atrial fibrillation, cardiac hypertrophy, and fibrosis \cite{11-15, new2}. The use of miRNAs as diagnostic and prognostic biomarkers in CVDs is supported by their stability and rapid release into circulation after myocardial injury \cite{11-15}. In CAD, altered expression of miRNAs like miR-1, miR-133a, miR-208a/b, and miR-499, which are abundantly expressed in the heart, has been reported in patients compared to healthy controls. Additional miRNAs including miR-21, miR-208a/b, miR-133a/b, and the miR-30 family are frequently dysregulated in acute coronary syndrome (ACS) versus stable CAD  \cite{new3}. Furthermore, miRNAs like miR-3113-5p, miR-223-3p, miR-499a-5p, and miR-133a-3p demonstrate potential as biomarkers to identify patients at risk of sudden cardiac death \cite{new4}. Moreover, miRNAs have shown diagnostic potential in other CVDs. For instance, miR-21 has been associated with cardiac injury and has been implicated in the pathology and recurrence of MI. Elevated levels of miR-21 have been observed in ACS patients and have been linked to cardiomyocyte apoptosis and cardiac hypertrophy. Similarly, miR-26 has been implicated in the pathology and recurrence of MI \cite{new5}. In addition to their diagnostic potential, miRNAs have also shown promise as prognostic biomarkers for adverse myocardial effects, sudden death, and risk assessment in MI and other CVDs. For example, miR-101 and miR-150 have been associated with flawed left ventricular contractility after MI, while miR-16 and miR-27a have been linked to an increased risk of adverse left ventricular remodeling \cite{11-15, new3}. These miRNAs may provide valuable prognostic information and aid in risk stratification for post-MI complications.

Numerous studies have investigated the potential of miRNAs as biomarkers for MI, revealing promising findings. For instance, miR-1 has been proposed as a potential biomarker for MI \cite{new3}. This miRNA has shown increased expression levels in patients with MI, suggesting its potential diagnostic value. Additionally, other miRNAs, such as miR-19b-3p, miR-208a, miR-223-3p, miR-483-5p, and miR-499a-5p, have demonstrated promising diagnostic accuracy for MI within a short time window after the onset of symptoms \cite{new4}. A recent systematic review compared the peak time and diagnostic accuracy of miRNAs and conventional biomarkers in MI. The results revealed miR-1-3p, miR-19b-3p, miR-208a, miR-223-3p, miR-483-5p, and miR-499a-5p had superior peak times within 4 hours and better accuracy versus cTn and Creatine kinase-MB, indicating their promise for early diagnosis. The strengths of miRNAs included their early peak expression, satisfactory sensitivity and specificity, and higher accuracy especially within the first few hours of symptom onset compared to conventional biomarkers \cite{sys}.
% Change E

It has been postulated that the function and diagnostic properties of miRNAs are beyond the myocardium in patients with CVD. Specifically, the expression of miRNAs can vary in different biofluids and cell components such as serum and peripheral blood mononuclear cells (PBMCs) \cite{36}. 
PBMCs are a fraction of white blood cells, including monocytes,
lymphocytes, macrophages, and other cells of the immune system \cite{PBMC-miR}. Emerging data indicate that PBMCs can be used as a valid
source of biomarkers for monitoring various pathological conditions. Of
note, the alteration of mRNAs and miRNAs under pathological conditions
provides valuable information about different kinds of disorders. PBMCs
can recapitulate the conditions of target tissues, thus providing a
highly sensitive and specific source of biomarkers \cite{Meder6}. Combined with these conditions, these cells are repositories of
dysregulated genes and miRNA expression profiles in CVDs \cite{PBMC-miR, Meder6}.

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO)
database (\url{https://www.ncbi.nlm.nih.gov/geo/}).
% Change S
To obtain robust classification performance between MI, healthy control, and CAD samples, sufficiently large sample sizes for each group are required. For this purpose, the GSE59867 dataset was selected, as it contains sizable numbers of both MI and CAD samples. To provide an equally large set of healthy controls, the GSE56609 and GSE54475 datasets containing healthy samples were also included. Combining these three datasets enabled comparative analysis between MI, CAD, and healthy control groups with adequate statistical power.
% Change E
All samples were produced using Affymetrix Human Gene 1.0 ST Array
platform (GPL6244). 
% Change S
This platform contains 189 miRNA probes based on the annotation data from the GEO database.
% Change E
Only healthy, CAD, and early-stage MI samples were
selected from these datasets for further analysis.
% Change S
Early-stage MI samples were analyzed to enable detection of miRNA biomarkers specific to the initial ischemia and infarction event, before extensive myocardial necrosis and remodeling occurs. Using samples from the early phase enhances identification of miRNA signals related to plaque rupture and MI onset versus stable CAD. Additionally, early-stage samples allow investigation of mechanisms initiating myocardial injury.
%Change E
The basic information
for the three datasets evaluated in this study is provided in Table
~\ref{tab:datasets}. Bioinformatics analyses 
%Change S
including  preprocessing, differential expression analysis, and functional and pathway enrichment
analyses
%Change E
were conducted using R, ver.
4.2.0 \cite{R}, and RStudio \cite{RStudio}. All plots and
graphics of these sections were created using the ggplot2 R package \cite{ggplot}.

@@

In the barcode algorithm, the normalized intensity of gene $i$ in
sample $j$ is denoted as $x_{ij}$. A user-defined parameter, $C$,
was introduced along with the standard deviation ($\sigma^{ne}$) and
mean ($\mu^{ne}$) of the non-expressed distribution. Based on these
values, the barcode representation of a sample was generated as a vector
consisting of ones and zeros. 
%Change S
The ones and zeros generated by the barcode algorithm refer to binary calls of whether or not a gene is estimated to be expressed (1) or not expressed (0) in each individual sample. 
%Change E
The barcode function within the R fRMA package was employed to implement the barcode
algorithm, utilizing the default value of $C$.

@@

% Change S
\subsubsection{Differentially expressed miRNAs}\label{differentially-expressed-miRNAs}

The differentially expressed miRNAs were defined as those miRNAs within the total DEGs (i.e. they had an FDR $< 0.05$ resulted from the Fisher's exact test comparing the sample groups).

% Change E

@@

Two different approaches were used to select miRNAs for model training.
The first approach was to use differentially expressed miRNAs.
% Change
To capture additional miRNAs with high discriminatory power for distinguishing MI from CAD despite not reaching differentially expression criteria, a secondary approach was used. miRNAs were selected based on having individual area under the receiver operating characteristic curves (AUC-ROCs) exceeding 0.8 for separating MI and CAD. This AUC-based approach identifies miRNAs with the best classification performance, unconstrained by statistical cutoffs. Using both the differentially expressed and AUC-based selection provides complementary methods to uncover miRNA biomarkers from both a biological and diagnostic perspective.
% Change.

\subsubsection{Differentially expressed miRNAs}\label{mirnas-in-degs}

In this approach, a two-layer architecture is deployed to the data to
maximize the prediction values. The first layer predicted whether a
sample was healthy or not, and the second layer separated MI from CAD in
the samples that were predicted as not healthy in the first layer. To
this end, a distinct ML model was trained for each layer. Because there
were a limited number of miRNAs in the DEGs, both layers were trained
with all of them. For further comparison with the models' performance,
the ROC curve of each miRNA for classifying healthy and not-healthy, as
well as CAD and MI, was generated using a logistic regression model.

@@

A support vector machine (SVM) model using RBF kernels was trained and
hypertuned using all miRNAs in the DEGs. 
% Change
To account for the substantial class imbalance between the healthy and not-healthy groups, with 51 samples in the minority healthy class compared to 157 combined CAD and MI samples, adjustments were made to the sample weights used during model training. Without compensating for the imbalance, the machine learning model would be biased towards the majority class and potentially ignore the minority class. To counteract this, the sample weights were empirically tuned, with the weight for healthy samples set to 1 and the weight for not-healthy samples set to 0.5. These values were determined through iterative testing to produce a model with strong performance on both classes despite the imbalance.
% Change
The ROC curve
and confusion matrix for the model are reported.

@@

\begin{table}
\centering
\caption{Total, up-, and down-regulated DEGs and differentially expressed miRNAs.}
\label{tab:DEGstab}
\begin{tabularx}{\textwidth}{XXXXX}
\hline
& Total DEGs & Up-regulated DEGs & Down-regulated DEGs & miRNAs\\
\hline
MI vs. Healthy & 860 & 323 & 537 & hsa-miR-186-5p, hsa-miR-21-3p, hsa-miR-32-3p\\
CAD vs. Healthy & 670 & 262 & 408 & hsa-miR-186-5p, hsa-miR-21-3p, hsa-miR-32-3p\\
MI vs. CAD & 260 & 144 & 116 & hsa-miR-186-5p\\
\hline
\end{tabularx}
\end{table}

@@

% Change S
\subsubsection{Differentially expressed miRNAs}\label{mirnas-in-degs}
Among the DEGs for MI/healthy and CAD/healthy comparison, hsa-miR-186-5p, hsa-miR-32-3p, and hsa-miR-21-3p were identified as differentially expressed miRNAs. The only differentially expressed miRNA in MI/CAD comparison was hsa-miR-186-5p (Table~\ref{tab:DEGstab}). The expression profiles of the three
miRNAs are shown in Figure~\ref{fig:Expall}.
% Change E

@@

\subsubsection{Differentially expressed miRNAs}\label{mirnas-in-degs-1}

The ROC curves of each miRNA in each layer are presented in Figure~\ref{fig:miRROC}. Using the
logistic regression model, the AUC-ROC values of hsa-miR-21-3p, hsa-miR-32-3p, and
hsa-miR-186-5p for separating healthy and not-healthy samples were 0.98, 0.99,
and 0.90, respectively (Figure~\ref{fig:miRROC}A). The accuracy of each
miRNA for classifying the samples into healthy and not-healthy groups on
the test set for hsa-miR-21-3p, hsa-miR-32-3p, and hsa-miR-186-5p was 0.92, 0.98, and 0.89,
respectively. The ROC curve of each miRNA for classifying MI and CAD
samples is presented in Figure~\ref{fig:miRROC}B. The AUC-ROC and
accuracy for hsa-miR-21-3p, hsa-miR-32-3p, and hsa-miR-186-5p in the test set were 0.85;
0.70; and 0.86, and 0.78; 0.67; and 0.74, respectively.

\begin{table}
\centering
\caption{Investigated miRNAs log fold-change and adjusted p-values for CAD samples relative to healthy, MI samples relative to healthy, and MI samples relative to CAD.}
\label{tab:mirExptable}
\begin{tabular}{lcccccc}
\toprule
\multicolumn{1}{c}{} & \multicolumn{2}{c}{CAD/Healthy} & \multicolumn{2}{c}{MI/Healthy} & \multicolumn{2}{c}{MI/CAD} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
& logFC & adj. p-value & logFC & adj. p-value & logFC & adj. p-value\\
\midrule
hsa-miR-186-5p & 1.4 & 3.60e-25 & 0.9 & 6.76e-20 & -0.5 & 1.05e-09\\
hsa-miR-21-3p & 1.4 & 1.31e-17 & 2.3 & 2.07e-47 & 0.8 & 2.96e-11\\
hsa-miR-32-3p & 2.5 & 8.39e-43 & 2.2 & 3.10e-59 & -0.3 & 7.60e-04\\
hsa-miR-197-5p & 0.5 & 2.95e-20 & 0.7 & 1.59e-47 & 0.2 & 8.58e-09\\
hsa-miR-29a-5p & 0.7 & 7.76e-29 & 0.1 & 1.70e-01 & -0.5 & 2.14e-10\\
hsa-miR-296-5p & -0.1 & 5.00e-02 & 0.1 & 2.00e-02 & 0.2 & 6.15e-06\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\centering 
\includegraphics[width=0.95\linewidth]{DEMs ROCs h not h cad mi}
\caption{ROC curve for the model trained with differentially expressed miRNAs on test set classification; (A) An SVM model with RBF kernel for healthy and not-healthy and (B) An SVM model with linear kernel for CAD and MI sample classification.}\label{fig:DEMROC}
\end{figure}

\begin{figure}
\centering 
\includegraphics[width=0.9\linewidth]{DEMs Conf. matrix}
\caption{Confusion matrix for the model trained with differentially expressed miRNAs on test set classification; (A) An SVM model with RBF kernel for healthy and not-healthy and (B) An SVM model with linear kernel for CAD and MI sample classification.}\label{fig:DEMCF}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.65\linewidth]{Models DEGs mirs}
\caption{Area under the receiver operating characteristic curve and accuracy of different models trained with three differentially expressed miRNAs.}\label{fig:DEMmodels}
\end{figure}

@@

After calculating the AUC-ROC for each miRNA to classify of MI and CAD
samples, the miRNAs with AUC-ROC $> 0.8$ were selected. The miRNAs
selected were hsa-miR-29a-5p, hsa-miR-197-5p, hsa-miR-186-5p, hsa-miR-21-3p, and hsa-miR-296-5p. The
expression levels of these miRNAs in healthy, CAD, and MI samples are
presented in Figure~\ref{fig:Expall}. The ROC curves of the selected
miRNAs in both layers are shown in Figure~\ref{fig:miRROC}.

@@

In this study, we combined three GEO datasets for healthy, CAD, and MI
samples. Having these sample sets alongside bioinformatics analysis and
ML methods enabled us to identify potential biomarker sets and effective
therapeutic targets. The results of the DEG analysis (Table
~\ref{tab:DEGstab} and Figure~\ref{fig:venn}) prove the close
relationship between the MI and CAD samples. Interestingly, functional
enrichment analysis demonstrated that DEGs in both CAD/healthy and
MI/healthy were strongly correlated with the immune cell response, which
is a major part of PBMCs. Two sets of miRNAs were selected as biomarker
sets for sample classification. Hsa-miR-21-3p; hsa-miR-32-3p; and hsa-miR-186-5p were
selected as differentially expressed miRNAs, and hsa-miR-186-5p; hsa-miR-21-3p;
hsa-miR-29a-5p; hsa-miR-197-5p; and hsa-miR-296-5p were selected based on their AUC-ROC
values. As shown in Figure~\ref{fig:miRROC}, all miRNAs selected with
both approaches had AUC-ROCs $> 0.9$ for isolating healthy and
not-healthy samples except for hsa-miR-296-5p and hsa-miR-29a-5p. The data confirmed
that the real challenge was to classify CAD and MI samples because of
the close overlap. Of the six miRNAs under investigation in both
approaches, except for hsa-miR-32-3p, all miRNAs had an AUC-ROC $> 0.8$ for
the discrimination of CAD and MI samples. As expected, the high AUC-ROC
values of the miRNAs confirmed their high potential as biomarkers.

@@

Numerous studies have reported that different biological processes can
affect the miRNA expression in PBMCs. However, the exact role of miRNAs
in the function of immune cells and the correlation between specific
pathological conditions and miRNA profiles remain controversial. Several
studies have proven the activation of particular miRNA types in PBMCs
under cardiovascular events \cite{296}. For instance, there is
evidence that elevation of hsa-miR-186-5p suppresses the expression of
cystathionine-$\gamma$-lyase, leading to the subsequent secretion of
pro-inflammatory cytokines and cellular lipid accumulation. In addition,
macrophage-derived hsa-miR-186-5p may promote atherosclerotic plaque formation
\cite{186}. In line with this claim, we found that hsa-miR-186-5p was
up-regulated in both CAD and MI candidates compared to their control
counterparts. Surprisingly, the obtained data indicated that the
expression of hsa-miR-186-5p was higher in patients with CAD than in patients
with MI (Figure~\ref{fig:Expall}). Specifically, hsa-miR-186-5p was the only
differentially expressed miRNA between CAD and MI, with a clear
up-regulation in CAD, indicating its main role in the promotion of
atherosclerosis.

As mentioned before, hsa-miR-21-3p was also up-regulated in both MI and CAD
patients compared to healthy controls. Moreover, the expression value of
hsa-miR-21-3p was significantly higher in the MI group than in the CAD group
(Table~\ref{tab:mirExptable}). It is thought that the up-regulation of
hsa-miR-21-3p in PBMCs is a compensatory reaction to reduce the T$_{reg}$
lymphocyte number in response to the reduction in TGF$\beta1$
secretion into the plasma through a TGF$\beta1$/smad-independent
pathway. In line with the previous and present data, hsa-miR-21-3p can modulate
the activity of PBMCs following the occurrence of cardiovascular
diseases \cite{21}.

>>>>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

@@

\author[2]{\fnm{Reza} \sur{Rahbarghazi}}

\author[3]{\fnm{Zeynab} \sur{Piryaei}}

\author[4]{\fnm{Mahdad} \sur{Esmaeili}}

\author[5]{\fnm{Çığır} \sur{Biray Avcı}}

@@

\author[2*]{\fnm{Kaveh} \sur{Kavousi}}\email{kkavousi@ut.ac.ir}

\affil[1]{\orgdiv{Department of Medical Nanotechnology,}, \orgname{Faculty of Advanced Medical Sciences}, \orgaddress{\street{Tabriz University of Medical Sciences}, \city{Tabriz}, \country{Iran}}}

\affil[2]{\orgdiv{Department of Applied Cell Sciences}, \orgname{Faculty of Advanced Medical Sciences}, \orgaddress{\street{Tabriz University of Medical Sciences}, \city{Tabriz}, \country{Iran}}}

\affil[3]{\orgdiv{Laboratory of Complex Biological Systems and Bioinformatics (CBB), Department of Bioinformatics}, \orgname{Institute of Biochemistry and Biophysics (IBB)}, \orgaddress{\street{University of Tehran}, \city{Tehran}, \country{Iran}}}

\affil[4]{\orgdiv{Medical Bioengineering Department}, \orgname{Faculty of Advanced Medical Sciences}, \orgaddress{\street{Tabriz University of Medical Sciences}, \city{Tabriz}, \country{Iran}}}

\affil[5]{\orgdiv{Medical Biology Department}, \orgname{School of Medicine}, \orgaddress{\street{Ege University}, \city{İzmir}, \country{Türkiye}}}

@@

\textbf{Results:} We identified miR-21, miR-186, and miR-32 as the only miRNAs among the
differentially expressed genes, and a set including miR-186, miR-21,
miR-197, miR-29A, and miR-296 as the optimum set of miRNAs selected by
their AUC-ROC. Both biomarker sets could distinguish healthy from
not-healthy samples with complete accuracy. The best performance for the
classification of CAD and MI was achieved with an SVM model trained
using the biomarker set selected by AUC-ROC, with an AUC-ROC of 0.96 and
an accuracy of 0.94 on the test data.

@@

Over the last 50 years, numerous attempts have been made to use
biomarkers to facilitate diagnosis, assess the risk, follow-up therapy,
and determine therapeutic efficacy in CVD candidates. Based on released
guidelines, cardiac troponins (cTns) are used as a highly sensitive and
accurate approach for detecting MI. Despite these inherent advantages,
the high sensitivity of cTn-based assays has also led to more
false-positive results \cite{17}, necessitating the advent
and development of new modalities with pathological value. To improve
the diagnostic value of existing MI biomarkers, a combination of
complementary biological markers, such as microRNAs (miRNAs) and other
genetic factors, has been proposed. Previous research supports the
notion that miRNAs exhibit great potential as alternative biomarkers for
CVD detection and follow-up \cite{miR}. It has been suggested
that miRNAs possess 18-22 nucleotides and play a crucial role in the
regulation of gene expression. Evidence indicates that miRNAs are
involved in the pathogenesis of cardiac tissue injury \cite{m1}. Several biological processes, such as angiogenesis,
cardiomyocyte growth and contractility, lipid metabolism, plaque
formation, and cardiac rhythm, are regulated by miRNAs \cite{m2}. These elements can easily circulate in biofluids and could be
considered theranostic targets in terms of CVDs \cite{m1}. It has been postulated that the function and diagnostic
properties of miRNAs are beyond the myocardium in patients with CVD.
Specifically, the expression of miRNAs can vary in different biofluids
and cell components such as serum and peripheral blood mononuclear cells
(PBMCs) \cite{36}.

PBMCs are a fraction of white blood cells, including monocytes,
lymphocytes, macrophages, and other cells of the immune system \cite{PBMC-miR}. Emerging data indicate that PBMCs can be used as a valid
source of biomarkers for monitoring various pathological conditions. Of
note, the alteration of mRNAs and miRNAs under pathological conditions
provides valuable information about different kinds of disorders. PBMCs
can recapitulate the conditions of target tissues, thus providing a
highly sensitive and specific source of biomarkers \cite{Meder6}. Combined with these conditions, these cells are repositories of
dysregulated genes and miRNA expression profiles in CVDs \cite{PBMC-miR, Meder6}.

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO)
database (\url{https://www.ncbi.nlm.nih.gov/geo/}). To obtain sufficient
classification power between MI, healthy, and CAD samples, a relatively
large sample size is needed Therefore, GSE59867 for the MI and CAD
samples and GSE56609 and GSE54475 for the healthy samples were selected.
All samples were produced using Affymetrix Human Gene 1.0 ST Array
platform (GPL6244). Only healthy, CAD, and early-stage MI samples were
selected from these datasets for further analysis. The basic information
for the three datasets evaluated in this study is provided in Table
~\ref{tab:datasets}. Bioinformatics analyses were conducted using R, ver.
4.2.0 \cite{R}, RStudio \cite{RStudio}. All plots and
graphics of these sections were created using the ggplot2 R package \cite{ggplot}.

@@

In the barcode algorithm, the normalized intensity of gene $i$ in
sample $j$ is denoted as $x_{ij}$. A user-defined parameter, $C$,
was introduced along with the standard deviation ($\sigma^{ne}$) and
mean ($\mu^{ne}$) of the non-expressed distribution. Based on these
values, the barcode representation of a sample was generated as a vector
consisting of ones and zeros, representing the estimated expression
(ones) and non-expression (zeros) of each gene. The barcode function
within the R fRMA package was employed to implement the barcode
algorithm, utilizing the default value of $C$.

@@

@@

Two different approaches were used to select miRNAs for model training.
The first approach was to use differentially expressed miRNAs. In the
second approach, miRNAs with individual area under the receiver operating characteristic curves (AUC-ROCs) over 0.8 for separating MI from CAD were selected. The results of these two approaches can
provide an informative comparison between the predictive capabilities of
sets of miRNAs selected with different logics.

\subsubsection{miRNAs in DEGs}\label{mirnas-in-degs}

In this approach, a two-layer architecture is deployed to the data to
maximize the prediction values. The first layer predicted whether a
sample was healthy or not, and the second layer separated MI from CAD in
the samples that were predicted as not healthy in the first layer. To
this end, a distinct ML model was trained for each layer. Because, there
were a limited number of miRNAs in the DEGs, both layers were trained
with all of them. For further comparison with the models' performance,
the ROC curve of each miRNA for classifying healthy and not-healthy, as
well as CAD and MI, was generated using a logistic regression model.

@@

A support vector machine (SVM) model using RBF kernels was trained and
hypertuned using all miRNAs in the DEGs. To handle the severe imbalance
in the number of samples (51 for the healthy group and 157 for the
not-healthy group), the sample weights for the healthy and the
not-healthy samples were set to 1 and 0.5, respectively. The ROC curve
and confusion matrix for the model are reported.

@@

\begin{table}
\centering
\caption{Total, up-, and down-regulated DEGs and differentially expressed miRNAs.}
\label{tab:DEGstab}
\begin{tabular*}{\textwidth}{@{\extracolsep\fill}lcccc}
\toprule
& Total DEGs & Up-regulated DEGs & Down-regulated DEGs & miRNAs\\
\midrule
MI vs. Healthy & 860 & 323 & 537 & miR-186, miR-21, miR-32\\
CAD vs. Healthy & 670 & 262 & 408 & miR-186, miR-21, miR-32\\
MI vs. CAD & 260 & 144 & 116 & miR-186\\
\bottomrule
\end{tabular*}
\end{table}

@@

@@

\subsubsection{miRNAs in DEGs}\label{mirnas-in-degs-1}

Among the DEGs, miR-186, miR-32, and miR-21 were identified as
differentially expressed miRNAs. The expression profiles of the three
miRNAs are shown in Figure~\ref{fig:Expall}. The ROC curves of each
miRNA in each layer are presented in Figure~\ref{fig:miRROC}. Using the
logistic regression model, the AUC-ROC values of miR-21, miR-32, and
miR-186 for separating healthy and not-healthy samples were 0.98, 0.99,
and 0.90, respectively (Figure~\ref{fig:miRROC}A). The accuracy of each
miRNA for classifying the samples into healthy and not-healthy groups on
the test set for miR-21, miR-32, and miR-186 was 0.92, 0.98, and 0.89,
respectively. The ROC curve of each miRNA for classifying MI and CAD
samples is presented in Figure~\ref{fig:miRROC}B. The AUC-ROC and
accuracy for miR-21, miR-32, and miR-186 in the test set were 0.85;
0.70; and 0.86, and 0.78; 0.67; and 0.74, respectively.

\begin{table}
\centering
\caption{Investigated miRNAs log fold-change and adjusted p-values for CAD samples relative to healthy, MI samples relative to healthy, and MI samples relative to CAD.}
\label{tab:mirExptable}
\begin{tabular}{lcccccc}
\toprule
\multicolumn{1}{c}{} & \multicolumn{2}{c}{CAD/Healthy} & \multicolumn{2}{c}{MI/Healthy} & \multicolumn{2}{c}{MI/CAD} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
& logFC & adj. p-value & logFC & adj. p-value & logFC & adj. p-value\\
\midrule
miR-186 & 1.4 & 3.60e-25 & 0.9 & 6.76e-20 & -0.5 & 1.05e-09\\
miR-21 & 1.4 & 1.31e-17 & 2.3 & 2.07e-47 & 0.8 & 2.96e-11\\
miR-32 & 2.5 & 8.39e-43 & 2.2 & 3.10e-59 & -0.3 & 7.60e-04\\
miR-197 & 0.5 & 2.95e-20 & 0.7 & 1.59e-47 & 0.2 & 8.58e-09\\
miR-29A & 0.7 & 7.76e-29 & 0.1 & 1.70e-01 & -0.5 & 2.14e-10\\
miR-296 & -0.1 & 5.00e-02 & 0.1 & 2.00e-02 & 0.2 & 6.15e-06\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\centering 
\includegraphics[width=0.95\linewidth]{DEMs ROCs h not h cad mi}
\caption{ROC curve for the model trained with miRNAs in DEGs on test set classification; (A) An SVM model with RBF kernel for healthy and not-healthy and (B) An SVM model with linear kernel for CAD and MI sample classification.}\label{fig:DEMROC}
\end{figure}

\begin{figure}
\centering 
\includegraphics[width=0.9\linewidth]{DEMs Conf. matrix}
\caption{Confusion matrix for the model trained with miRNAs in DEGs on test set classification; (A) An SVM model with RBF kernel for healthy and not-healthy and (B) An SVM model with linear kernel for CAD and MI sample classification.}\label{fig:DEMCF}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.65\linewidth]{Models DEGs mirs}
\caption{Area under the receiver operating characteristic curve and accuracy of different models trained with three miRNAs in DEGs.}\label{fig:DEMmodels}
\end{figure}

@@

After calculating the AUC-ROC for each miRNA to classify of MI and CAD
samples, the miRNAs with AUC-ROC $> 0.8$ were selected. The miRNAs
selected were miR-29a, miR-197, miR-186, miR-21, and miR-296. The
expression levels of these miRNAs in healthy, CAD, and MI samples are
presented in Figure~\ref{fig:Expall}. The ROC curves of the selected
miRNAs in both layers are shown in Figure~\ref{fig:miRROC}.

@@

In this study, we combined three GEO datasets for healthy, CAD, and MI
samples. Having these sample sets alongside bioinformatics analysis and
ML methods enabled us to identify potential biomarker sets and effective
therapeutic targets. The results of the DEG analysis (Table
~\ref{tab:DEGstab} and Figure~\ref{fig:venn}) prove the close
relationship between the MI and CAD samples. Interestingly, functional
enrichment analysis demonstrated that DEGs in both CAD/healthy and
MI/healthy were strongly correlated with the immune cell response, which
is a major part of PBMCs. Two sets of miRNAs were selected as biomarker
sets for sample classification. miR-21; miR-32; and miR-186 were
selected as differentially expressed miRNAs, and miR-186; miR-21;
miR-29a; miR-197; and miR-296 were selected based on their AUC-ROC
values. As shown in Figure~\ref{fig:miRROC}, all miRNAs selected with
both approaches had AUC-ROCs $> 0.9$ for isolating healthy and
not-healthy samples except for miR-296 and miR-29a. The data confirmed
that the real challenge was to classify CAD and MI samples because of
the close overlap. Of the six miRNAs under investigation in both
approaches, except for miR-32, all miRNAs had an AUC-ROC $> 0.8$ for
the discrimination of CAD and MI samples. As expected, the high AUC-ROC
values of the miRNAs confirmed their high potential as biomarkers.

@@

Numerous studies have reported that different biological processes can
affect the miRNA expression in PBMCs. However, the exact role of miRNAs
in the function of immune cells and the correlation between specific
pathological conditions and miRNA profiles remain controversial. Several
studies have proven the activation of particular miRNA types in PBMCs
under cardiovascular events \cite{296}. For instance, there is
evidence that elevation of miR-186 suppresses the expression of
cystathionine-$\gamma$-lyase, leading to the subsequent secretion of
pro-inflammatory cytokines and cellular lipid accumulation. In addition,
macrophage-derived miR-186 may promote atherosclerotic plaque formation
\cite{186}. In line with this claim, we found that miR-186 was
up-regulated in both CAD and MI candidates compared to their control
counterparts. Surprisingly, the obtained data indicated that the
expression of miR-186 was higher in patients with CAD than in patients
with MI (Figure~\ref{fig:Expall}). Specifically, miR-186 was the only
differentially expressed miRNA between CAD and MI, with a clear
up-regulation in CAD, indicating its main role in the promotion of
atherosclerosis.

As mentioned before, miR-21 was also up-regulated in both MI and CAD
patients compared to healthy controls. Moreover, the expression value of
miR-21 was significantly higher in the MI group than in the CAD group
(Table~\ref{tab:mirExptable}). It is thought that the up-regulation of
miR-21 in PBMCs is a compensatory reaction to reduce the T$_{reg}$
lymphocyte number in response to the reduction in TGF$\beta1$
secretion into the plasma through a TGF$\beta1$/smad-independent
pathway. In line with the previous and present data, miR-21 can modulate
the activity of PBMCs following the occurrence of cardiovascular
diseases \cite{21}.

Recent data have supported the elevation of miR-32 levels in CAD samples
with calcification of the coronary artery. Notably, miR-32 promotes
vascular smooth muscle calcification in mice by controlling the activity
of several proteins, including bone morphogenetic protein-1,
runt-related transcription factor-2 (RUNX2), osteopontin, and
bone-specific phosphoprotein matrix GLA protein \cite{32-2}.
Likewise, some reports are associated with the activity of miR-32 in
PBMCs in several pathologies \cite{magic, 32}. The
exact role of miR-32 in PBMCs after cardiovascular events remains
unclear.
